Revisiting the Nos\'{e}-Hoover chain thermostat applied to the Rigid Bodies Molecular Dynamics

### 2017-02-17 rigid_bodies.tex
aa

\title{A Simplified Formulation for Molecular Dynamics with Rigid Bodies}

@@

Molecular Simulation is the computacional realization of Statistical Mechanics\cite{Tuckerman2010} and requires numerical approximations in order to make possible a microsopic insight into complex physical phenomena. Under those approximations, certain Statistical Mechanics laws might not been necessarily traduced into assumptions in Molecular Simulation. For instance, it has already been shown that the unavoidable truncation errors associated with the numerical integrations of the equations of motion in Molecular Dynamics (MD) disrupt the equipartition of energy, and as a consequence the temperature becomes an undefined property\cite{Eastwood_2010}. In a previous paper,\citep{Abreu_2017} we observed the same problem now regarding the partition of the kinetic energy between translational and rotational degrees of freedom in MD of rigid bodies in the microcanonical (NVE) ensemble. That finding brough into question the method introduced by Kamberaj and co-workers\cite{Kamberaj2005}, which couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom and fails to reproduce the specified temperature for certain time step sizes.

The importance of rigid-body Molecular Dynamics stems from the fact that it is not limited to simulate molecules modelled as rigid bodies only, like the ubiquitous case of water,  but also because it is a coarse-grained strategy used for sistems designed as collections of rigid bodies like proteins and nanoparticiles\cite{Knorowski2012, Patra2013}. The approach of Kamberaj~\textit{et al.}\cite{Kamberaj2005}, which is implemented in the LAMMPS software package\cite{Plimpton1995}, has actually been used in the simulation of small molecules\cite{Aimoli_2014}, as well as in molecular motors\cite{Yan_2008,Akimov_2012,Geiger_2013}.

The Nos\'{e}-Hoover thermostat, routinely used in Molecular Dynamics, relies on the ``extended'' phase space on which ficticius variables are introduced in order to control the fluctuations of kinetic energy of the physical variables maintaining the temperature of the physical system in its specified value. The dynamics of those variables introduces non-Hamiltonian components to the equations of motion, which makes its evaluation more limited, comparing with Hamiltonian dynamics where backward error analysis is possible.

In this paper, we focus on the formulation of the equations of motion for rigid bodies in the canonical ensemble, and introduce time reversible and measure-preserving numerical integrators as well. both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat. Our starting point to develop the thermostatized equations of motion begining from our previous work on the microcanonical ensemble.

In the Nos\'{e}-Hoover thermostat we will carried out a heuristic analysis comparing with results obtained with Hybrid Monte Carlo (HMC)\cite{Duane1987}, which is an exact method in the sense that the truncation erros are not present, point of reference. We also using SHAKE\cite{Ryckaert1977}, which is an iterative method commonly used in MD simulations, which employes lagrange multiplier to treat the holonomic constraints used in MD simulations. For molecules modeled as rigid structures the dynamics of rigid bodies emerges as a suitable approach. This is expected to be computationally more efficient because the constraints are treated implicitly. The property of being symplectic implies the conservation of volume in phase space. the inverse statement is not true. measure-preserving

The paper is organized as follows. In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVT dynamics, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}. Moreover, we analyze the statistical distributions of the sampled energies in the NVT case, and check the equation we developed for computing pressure. Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.

\section{Theory}

Prompted by a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we have reformulated the Hamiltonian equations of motion for rigid bodies, from which we obtained symplectic integrators in the microcanonical (NVE) ensemble. In the aforementioned paper, we introduced an analytical solution for torque-free rotations, which would be particularly useful if any backward error analysis\cite{Hairer_2006} is to be carried out. However, the  Nos\'{e}-Hoover chain thermostat introduces non-Hamiltonian components, so any backward error analysis it is not feasible anymore. Given that, in this paper, the thermostatted version of the equations of motion will be generated considering the approximate solution for the rotational motion, which is also a mathematical simplification of the formulation introduced by Miller~\textit{et al.}\cite{Miller2002}, and exhibits a precision notably similar to that achieved using the analytical solution. In the next subsection, we will skip over the mathematical details and restrict ourselves to some fundamental expressions on which the dynamics in the canonical (NVT) ensemble will be developed. For full details in the NVE ensemble the reader is referred to our previous paper.\cite{Abreu_2017}

>>>>

\title{Revisiting the Nos\'{e}-Hoover chain thermostat applied to the Rigid Bodies Molecular Dynamics}

@@

Molecular Simulation is the computacional realization of Statistical Mechanics\cite{Tuckerman2010} that makes possible a microscopic insight into complex physical phenomena. To mimic ``real'' systems, numerical approximation are required under which certain Statistical Mechanics laws cannot be assumed to be valid also in Molecular Simulation. For instance, it has already been shown that the unavoidable truncation errors associated with the numerical integration of the equations of motion in Molecular Dynamics disrupt the equipartition of energy, and as a consequence the temperature becomes an undefined property\cite{Eastwood_2010}. In a previous paper,\citep{Abreu_2017} we observed the same problem, but regarding the partition of the kinetic energy between translational and rotational degrees of freedom in Molecular Dynamics of rigid bodies in the microcanonical ensemble. That finding brough into question the method introduced by Kamberaj and co-workers\cite{Kamberaj2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom and fails to reproduce the specified temperature for certain time step sizes.

The relevance of rigid-body Molecular Dynamics stems from the fact that its application is not limited to molecules actually modelled as a rigid body, like the ubiquitous case of water, but also it constitues a simple coarse-grain strategy for systems that might be designed as collection of rigid bodies interconected, like proteins and nanoparticules\cite{Knorowski2012, Patra2013}. As a matter of fact, the approach of Kamberaj~\textit{et al.}\cite{Kamberaj2005}, which is implemented in the LAMMPS software package\cite{Plimpton1995}, has been applied in the simulation of small molecules\cite{Geiger_2013,Aimoli_2014,Aimoli_2014_2}, as well as in membranes~\cite{Bucior_2012} molecular motors\cite{Akimov_2012}, micelles\cite{Yan_2008} and nanoparticles\cite{Patra_2014}, DNA\cite{Fathizadeh_2012}

The Nos\'{e}-Hoover chain thermostat\cite{Martyna1992}, routinely used in Molecular Dynamics, employs the extended phase-space approach and introduces ficticius variables in order to ``control'' the kinetic energy fluctuations of the physical variables which allows to maintain the system at a specified temperature. The dynamics of those variables introduces non-Hamiltonian components which makes infeasible to carry out the backward error analysis. Altough it is not possible to devise symplectic integrator, one can rigurously obtain measure-preserving numerical solvers\cite{Sergi2001,Ezra2004,Ezra2006} and obtain numerical stability.

In this paper, we focus on the formulation of the equations of motion for rigid bodies in the canonical ensemble and introduce time reversible and measure-preserving numerical integrators as well. Our startint point is the symplectic integrator in the microcanonical ensemble we introduced in a previous paper\citep{Abreu_2017}, which employs the unit quaterion representation for the rotational degrees of freedom. Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at the specified value. The analysis of the thermostats will be rather heuristic reliying on comparisons with results obtained employing the Hybrid Monte Carlo (HMC)\cite{Duane1987}, which is an exact method in the sense that the truncation erros are not present. We will also include results from the SHAKE\cite{Ryckaert1977} method, which is an iterative method commonly used in MD simulations, which employes lagrange multiplier to treat the holonomic constraints. unavoidable errors but we shoud be aware of the effect on the properties of interest.

The paper is organized as follows. In Sec.~\ref{sec:theory} we briefly review the equations of motion in the microcanonical ensemble, which are the basis for developing the corresponding dynamics in the canonical ensemble. In Sec.~\ref{sec:numerical_solvers} we devise the corresponding numerical integrators, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}. Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.

\section{Theory}
\label{sec:theory}
Prompted by a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we have reformulated the Hamiltonian equations of motion for rigid bodies, from which we obtained symplectic integrators in the microcanonical (NVE) ensemble. In the aforementioned paper, we introduced an analytical solution for torque-free rotations, which would be particularly useful if any backward error analysis\cite{Hairer_2006} is to be carried out. However, the  Nos\'{e}-Hoover chain thermostat introduces non-Hamiltonian components, so any backward error analysis it is not feasible anymore. Given that, in this paper, the thermostatted version of the equations of motion will be generated considering the approximate solution for the rotational motion, which is also a mathematical simplification of the formulation introduced by Miller~\textit{et al.}\cite{Miller2002}, and exhibits a precision notably similar to that achieved using the analytical solution. In the next subsection, we will skip over the mathematical details and restrict ourselves to some fundamental expressions on which the dynamics in the canonical (NVT) ensemble will be developed. For full details in the NVE ensemble the reader is referred to our previous paper.\cite{Abreu_2017}
